\thetitle

### 2008-05-29 paper.tex
Another one bites the dust...

More than twenty years of research in analysis and modeling of network traffic have yielded tremendous advances in our understanding of the complex behaviors that emerge from the interaction of millions of humans and computers on the Internet and in local area networks~(\caps{LAN}s).
Both traffic analysis and realistic workload generation, however, remain active areas of research with many unanswered questions.
One of the major challenges of modeling full-network traffic is the multilevel interdependence of packet, flow\flownote and node behaviors.
Three types of simplifying assumptions are commonly applied for both analysis and generation:
\begin{enumerate}
\item \textbf{Uniformity:} that certain aspects of traffic behavior are uniformly distributed. For example, that  all nodes in a network have equal probability of being the source or destination of packets or flows.
\item \textbf{Independence:} that traffic properties share a single marginal\marginalnote distribution across the entire network, and individual values are independently drawn from this marginal distribution. For example, assuming that a single distribution describes an entire network, as opposed to modeling each flow with its own distribution of packet sizes.
\item \textbf{Subdivided Independence:} that traffic properties behave independently, but only within certain groupings. The assumption, for example, that all flows emanating from a node share the same distribution of packet sizes.
\end{enumerate}
We would be arguing against a straw man if we implied that researchers who apply these assumptions actually believe them to be true.
They clearly are not: flows have drastically different packet size distributions; different nodes initiate and receive drastically different numbers of packets and flows. Consider downloading a large file vs. typing in an \caps{SSH} session, and BitTorrent users vs. occasional email checkers.

@@

The second reason why uniformity and independence assumptions are so commonly used is that there is no generally accepted alternative for analyzing traffic.
The difficulty lies in the high dimensional and stochastic nature of network behaviors.
Suppose that we assume that packet sizes and inter-packet intervals are drawn from some distribution for each flow.
Prior research has indicated that, given accurate per-flow distributions, this is a realistic generative model of network behavior~\cite{Karpinski07:realism}.\realismnote
However, such empirical distributions are complex entities unto themselves---and worse still, each flow has its own distribution.
Modeling the entire network's behavior thus entails producing a distribution of distributions---with only one sample value from each distribution.
% How can we know if two flows' observed behaviors differ because they were fundamentally different or because the stochastic processes driving them happened to produce different-looking results?

Uniform and independent modeling collapses network behavior (in a way that we will make precise later) to drastically reduce the dimensionality of the representation.
This observation yields two immediate questions:
\begin{enumerate}
\item How can we represent the original ``uncollapsed'' network behavior?
\item Is there a better way than assuming independence or uniformity to reduce dimensionality?
\end{enumerate}
It is already common practice in network modeling to think of flow behaviors as vectors.
The \caps{IP} protocol number, source and destination \caps{IP} addresses, and \caps{UDP/TCP} source and port numbers that define a packet flow are commonly referred to as a 5-tuple---i.e., a vector in five dimensions.
Other aspects of a flow's behavior can be represented vectorially as well, but there is a one subtlety that may not be immediately obvious.

By definition, vectors can be composed via addition, yielding a new vector representing the aggregate of the two constituent vectors.
If we are to represent flow behaviors as vectors, then we must ensure that our representation respects addition of vectors: the sum of two flow representations should represent the aggregated behavior of the two flows.
We demonstrate how this can be accomplished and argue that the natural generalization of this approach is to represent an entire network's behavior as a large matrix, with each row representing a single flow.
We call such a representation of flow behavior a \emph{linear representation} since it possesses the fundamental characteristic of linearity:
\begin{align}
  f(x + y) = f(x) + f(y).
\end{align}

The matrix representation gives an uncollapsed representation of network behavior that preserves the non-uniformity and non-independence of flow behaviors.

@@

% The question remains whether assumptions of uniformity and independence are benign or whether they severely distort the properties of network behavior we are trying to analyze or replicate experimentally.
% %Is the analysis and generation of network traffic significantly distorted by such assumptions?
% This is the first fundamental question that this paper addresses. The that uniformity assumptions has already been made in earlier papers~\cite{Karpinski07:cbr-failure}.
% Therefore, we focus here on addressing assumptions of independence or subdivided independence. We find that assuming independent behaviors across entire networks is also extremely distorting of important measures of network performance.
% %This research has shown that assumptions of constant bit-rate (\caps{CBR}) flow behavior, as well as assumptions of uniformity at the flow and node levels of behavior drastically distort important performance metrics when used to generate wireless workload traffic for experiments.
% 
% While these assumptions certainly simplify both analysis and workload generation, they are quite certainly not true: flows do not all share a single packet size distribution; every node is not equally likely to be the source or destination of each packet or flow.
% 
% % The rate of packet arrivals at routers on the Internet backbone is known to exhibit complex, self-similar patterns with long-range dependence~\cite{Paxson95,Paxson96,Willinger98}. Other research has demonstrated how to accurately model the behavior of a single flow of packets between two application endpoints, including ``closed-loop'' feedback induced by the ubiquitous transmission control protocol~(\caps{TCP})~\cite{Hernandez06:dissertation}. Recently, the high-level statistical properties of wireless \caps{LAN} traffic have been rigorously modeled with a variety of parametric distributions~\cite{Hernandez06:wlan-traffic}.
% 
% Despite great breakthroughs in understanding and methodology, the networking community remains without a ``cookbook'' for generating realistic \caps{LAN} workloads. By ``cookbook,'' we mean an explicit method for generating network traffic---down to the size and transmission time of each application data unit. This lack is particularly relevant to wireless research, since the medium is fundamentally resource-limited, and consequently local traffic patterns can have a huge impact on overall network performance~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
% 
% In the current standard practice for experimental workload generation, drastic assumptions of uniformity abound. Flows are commonly modeled as constant bit-rate (\caps{CBR}) streams of identical packets sent at an invariant rate. The flows in an experiment are commonly given the same duration and data rate. Nodes in the network are often selected as flow endpoints randomly or arbitrarily. While more complex approaches have been proposed~\cite{Sommers04,Avallone04,Avallone06}, % TODO: better references here.
% the simplest, most na\"ive models remain standard because no objective criterion exists by which to compellingly argue that any one model is better than the rest. It is simply unclear that simplistic models do any harm, or how, if at all, complex models provide benefit.
% 
% We believe that two major breakthroughs are needed to improve traffic modeling. The first requirement is an objective standard of realism by which to compare traffic models. Such a standard has already been introduced and used to demonstrate that uniform traffic modeling, including the ubiquitous \caps{CBR} model, drastically distort performance metrics at all levels of the network protocol stack~\cite{Karpinski07:realism,Karpinski07:cbr-failure}. This standard serves to demonstrate the harm of overly simplistic models, such as \caps{CBR}, and provides the means to test whether more complex models deliver the desired improvements.
% The second breakthrough that is needed to improve modeling is a unified framework for expressing a broad variety of traffic models, that allows researchers to systematically express and reason about the assumptions and simplifications entailed in each model.
% 
% The major contribution of this paper is addressing this second requirement. We propose the General Matrix Model (\caps{GMM}) for local network traffic.  In this model, the traffic pattern of an entire network is represented as a large, sparse matrix. Each row vector represents a single flow of packets in the network, with components representing the \caps{IP} protocol type, source and destination nodes, flow size, start time, and packet behavior.  The crucial property of this representation is that it respects vector space operations: summation of behavior vectors gives an accurate description of the aggregate behavior of the flows represented by those vectors. One benefit of this representation is that it makes it immediately possible to express a broad variety of assumptions and conditions on traffic patterns in terms of simple matrix operations. It also provides vital mathematical clarity and allows us to apply the theoretical and computational tools of modern linear algebra to the problems of traffic analysis and modeling.
% 
% After presenting \caps{GMM} and discussing how common concepts in traffic modeling can be expressed in precise mathematical terms within this framework, we proceed with experimental validation of \caps{GMM}. We use the technique of \textit{paired differential simulation} that the authors proposed and used in previous work~\cite{Karpinski07:realism,Karpinski07:cbr-failure}. This technique uses real wireless trace traffic as an experimental control and conducts paired experiments comparing network performance from simulations using control traffic against others using synthetically generated traffic. The results are very positive: \caps{GMM} accurately reproduces network performance characteristics at all levels of the protocol stack.

%This does not solve the ``cookbook'' problem, however, because the \caps{GMM} must still be based on real traffic in order to reproduce its behavioral characteristics. It does however, reduce the problem of general traffic modeling to a more constrained problem of matrix modeling.

%Our experiments with various common assumptions of uniformity and regularity in network behavior yield mixed conclusions: most straightforward assumptions of homogenous behavior distort performance results significantly, but there are notable exceptions. The clearest exception is the time domain: traffic models appear to be relatively insensitive to local perturbation of the start time of flows---so long as the overall distribution remains approximately correct. Our conclusion is that traffic modeling needs to discard its assumptions of uniformity, regularity and independence. Instead we need to move ahead trying to understand the complex dependencies between the many aspects of traffic behaviors. By reducing traffic modeling to the well-understood domain of numerical linear analysis, the General Matrix Model brings us significantly closer to the ultimate goal of having a ``cookbook'' model for generating arbitrary, realistic experimental workload patterns.

%Worse still, it is unclear what ``realistic'' should mean in this context. Foremost, a clear and testable definition of realism for traffic models is needed. In the absence of such a definition, it remains standard in wireless experiments to generate traffic using the simplest possible model: all nodes, flows and packets are assumed to behave uniformly. Flows are commonly modeled as constant bit-rate (\caps{CBR}) streams of identical packets sent at an invariant rate; the flows in an experiment are commonly given the same duration and data rate; nodes in the network are often selected as flow endpoints randomly or arbitrarily. While more complex approaches have been proposed~\cite{Sommers04,Avallone04,Avallone06}, % TODO: better references here.
%the simplest, most na\"ive models remain standard because no objective criterion exists by which to compellingly argue that other models are any better. It is simply unclear that simplistic models do any harm, or how, if at all, complex models provide benefit.

%To address this foundational problem in network traffic modeling, we have introduced a fundamentally different approach to evaluating the realism of workload models~\cite{Karpinski07:realism,Karpinski07:cbr-failure}. Rather than subjectively choosing statistical measures that may or may not actually influence network performance, the realism of a model is measured directly in terms of its ability to accurately reproduce critical performance metrics. We define a traffic model to be \textit{sufficiently realistic} with respect to a given performance metric, if the model produces metric values similar to those observed using the original trace to generate traffic. Using this new approach, we explore the space of traffic models, establishing rigorously that the commonly used uniform models for node, flow and packet behavior drastically misrepresent important \caps{WLAN} performance metrics at all levels of the protocol stack.

%It is natural to compare these common, na\"ive models with the high-level parametric models proposed by \FHC~\cite{Hernandez06:wlan-traffic}. Here, however, we face another fundamental challenge in traffic modeling. Even with specific models for various aspects of network behavior, it remains non-trivial to combine them into a coherent composite model capable of generating concrete workloads. The {\FHC} models, for example, specify the distribution of flows per node in the network; since every flow has two endpoints, however, it remains unclear how to assign endpoints to flows such that the marginal distribution of flow per node matches the BiPareto model indicated. Similarly, the models dictate BiPareto marginal distribution of flow sizes across the entire network, but say nothing about how these sizes are to be distributed among the nodes. We could assume that the number of flows per node is independent from the size of the flows at that node, but, as we shall see, this assumption is invalid. Moreover, the details of flow behavior---packet sizes and inter-packet intervals---remain entirely undefined.

%Our point is not to denigrate the models proposed in~\cite{Hernandez06:wlan-traffic} in any way. On the contrary: these parametric models are the state of the art in modeling traffic patterns for wireless local area networks. The purpose, rather, is to point out how large a gap still remains between having accurate high-level models of marginal aspects of network behavior and being able to generate usable workloads with those models. Moreover, a great deal of ambiguity lies in that gap since there is no single obvious way to combine models.

%To clarify, disambiguate and standardize the specification of generative local area traffic models, we propose the \textit{Traffic Object Model} (\textit{\caps{TOM}}). The \caps{TOM} entails a hierarchy of abstract classes: \class{Node}, \class{Flow}, and \class{Packet} are the primary classes; the \class{Network} class collects everything. Each abstract class defines an interface of properties. A traffic model defines concrete classes at each level that provide the required interfaces; each class defines its own properties by delegation, inheritance, or overriding properties of other classes. Because of the separation of the \caps{TOM} into well-defined levels, each concrete class naturally defines a sub-model, specific to that level's behavior; sub-models can be mixed and matched combinatorially.

%We define a range of sub-models at each level of the \caps{TOM}, with decreasing similarity to the original trace behavior: \class{Trace} (reproduce original behavior), \class{Nonparametric}, \class{Parametric}, and \class{Uniform}. The control model uses the Trace sub-model at every level, exactly reproducing the behavior recorded from the original network. The standard na\"ive model uses the \class{Uniform} sub-model at all levels. The models proposed by {\FHC} are implemented by applying the \class{Parametric} sub-models at all levels. The \class{Nonparametric} sub-models provide a measure of how much error is introduced by parametric approximation and parameter estimation.

% objects and their properties together with rules for delegating, inheriting, and overriding the properties across the objects in the \caps{TOM} hierarchy. The \caps{TOM} allows

%The {\FHC} models, like most network behavior models, are marginal, meaning that they are fitted to data aggregated of over all nodes in the network, and not fitted on a per-node basis. With marginal models, since all information about interactions between the models is lost, so we must apply each model independently and hope that the result is good enough.

%Before we can define traffic models, however, we face the next major fundamental challenge in traffic modeling.

%In this paper, we address this fundamental problem in network traffic modeling by presenting an 
%In this paper, we propose objective criteria by which to judge whether

%The experimental evaluation of new wireless technologies inevitably requires network traffic. The most commonly used models for generating wireless local-area network (\caps{WLAN}) workload are simple, treating both nodes and flows uniformly, and are generally understood to be na\"ive. They remain popular because of the lack of compelling alternatives. The scope of impact of different workloads on network performance is largely unresearched and thus, poorly understood. We believe that one of the major handicaps preventing wireless experiments from more accurately predicting real-world performance is the lack of realistic models of \caps{WLAN} workload.

%The evaluation of wireless technology requires the generation of workload to test the viability and performance of the new protocol or technique being studied. We believe that lack of realism in traffic workload generation is one of the major limiting factors that prevents simulations and experimental test-bed deployments from accurately predicting the real-world performance of wireless technologies. Today, very little is understood about the impact of different workloads on network performance. Uniform constant bit-rate traffic (\caps{CBR}) is commonly used to evaluate protocols, but there is no evidence that behavior under such workloads is an accurate predictor of performance under real usage patterns. The inability to experimentally forecast real-world performance is a severe handicap to the entire networking community. It hinders the ability to effectively develop better solutions to the many difficult problems that face emerging wireless technologies.

%One of the hardest problems in modeling phenomena as complex as network traffic patterns lies in determining which characteristics of any given exemplar of behavior are essential and which can safely be abstracted away. In \caps{WLAN} traffic modeling, important characteristics have largely been determined based on intuition. In previous work~\cite{Karpinski07:realism}, we introduced a fundamentally different approach to evaluating the realism of workload models. Rather than subjectively choosing statistical measures that may or may not actually influence network performance, the realism of a model is measured directly in terms of its ability to accurately reproduce critical performance metrics. We define a traffic model to be \textit{sufficiently realistic} with respect to a given performance metric, if the model produces metric values that are similar to those observed using the original trace to generate traffic. Using this new approach, we explore the space of traffic models, establishing rigorously for the first time that common but simplistic models of \caps{WLAN} traffic drastically misrepresent important performance metrics at all levels of the protocol stack.

%Our analysis shows that the standard constant bit-rate (\caps{CBR}) traffic model, with uniformly chosen nodes and uniform flows of traffic, significantly misrepresents important performance metrics at every level of the network. To qualify these results in more depth, we also investigate partially synthetic variants of this standard model, allowing us to reason about which aspects of workload models have the most impact. For the most commonly used model, the delay of application data traversing the network from sender to receiver is underestimated by more than a factor of 5, \textit{on average}, and by more than a factor of 11 in a quarter of usage scenarios. Network control overhead is overestimated, on average, by a factor of about 2.5, while link control overhead is underestimated by almost a factor of 2. Even for metrics where the average and median misrepresentations are not so extreme, the statistical characteristics of error values often indicate that something is fundamentally unrealistic. None of the partially synthetic models manage to accurately represent more than one or two performance characteristics. We further use the \caps{AODV}~\cite{rfc:aodv} and \caps{OLSR}~\cite{rfc:olsr} ad~hoc routing protocols to show that the relative performance of protocols can be inverted when changing from using a simplistic traffic model to using real traffic: using random endpoint, uniform \caps{CBR} traffic, \caps{AODV} appears to induce less link-layer overhead than \caps{OLSR}, whereas using real traffic, it in fact induces more.

%The first major contribution of this work is the definition of ``sufficient realism,'' together with the analytical and statistical methodology to rigorously test whether synthetic traffic models meet this definition. The second contribution is the conclusion that the most commonly used model of wireless traffic drastically and consistently misrepresents some of the most important metrics for performance evaluation of wireless protocols. Many performance comparisons based on this model may need revisiting. The last and most important contribution is the collection of ideas and analytical tools necessary to create more realistic synthetic traffic models in the future. By applying the methods developed in this work, it will be possible to discover precisely which aspects of network usage affect the realism of performance results. Once this is known, it will become feasible to create models that accurately reproduce those aspects of real wireless usage.

>>>>

More than twenty years of research in analysis and modeling of network traffic have yielded tremendous advances in our understanding of the complex behaviors that emerge from the interaction of millions of humans and computers on the Internet and in packet-switched networks.
Both traffic analysis and realistic workload generation, however, remain active areas of research with many unanswered questions.
One of the major challenges of modeling full-network traffic is the multilevel interdependence of packet, flow\flownote and node behaviors.
Three types of simplifying assumptions are commonly applied in both analysis and generation:
\begin{enumerate}
\item \textbf{Uniformity:} that certain aspects of traffic behavior are uniformly distributed. For example, that  all nodes in a network have equal probability of being the source or destination of packets or flows.
\item \textbf{Independence:} that traffic properties share a single marginal\marginalnote distribution across the entire network, and individual values are independently drawn from this marginal distribution. For example, assuming that a single distribution describes an entire network, as opposed to modeling each flow with its own distribution of packet sizes.
\item \textbf{Subdivided Independence:} that traffic properties behave independently, but only within certain groupings. The assumption, for example, that all flows emanating from a node share the same distribution of packet sizes is of this type.
\end{enumerate}
We would be arguing against a straw man if we implied that researchers who apply these assumptions actually believe them to be true.
They clearly are not: flows have drastically different packet size distributions; different nodes initiate and receive drastically different numbers of packets and flows.
Simple examples illustrate this: downloading a large file vs. typing in an \caps{SSH} session; BitTorrent users vs. occasional email checkers.

@@

The second reason why uniformity and independence assumptions are so commonly used is that there are no generally accepted alternatives for representing or analyzing traffic.
The difficulty lies in the high dimensional and stochastic nature of network behaviors.
Suppose that we assume that packet sizes and inter-packet intervals are drawn from some distribution for each flow.
Prior research has indicated that, given accurate per-flow distributions, this is a fairly realistic model of network behavior~\cite{Karpinski07:realism}.\realismnote
Such empirical distributions, however, are complex entities unto themselves---and worse still, each flow has its own distribution.
Modeling the entire network's behavior thus entails producing a distribution of distributions.
% How can we know if two flows' observed behaviors differ because they were fundamentally different or because the stochastic processes driving them happened to produce different-looking results?

Uniform and independent modeling reduce the dimensionality of network behavior drastically, making analysis and generation tractable problems.
There are two immediate questions:
\begin{enumerate}
\item Can we represent the original ``unreduced'' network behavior in such a way that analysis and generation are still tractable problems?
\item Is there a better way to reduce dimensionality than assuming independence or uniformity of behaviors?
\end{enumerate}
The second contribution of this paper is to answer these questions and to propose a mathematical representation of network traffic that preserves the non-uniformity and non-independence of flow behaviors in a network, but makes traffic analysis and workload generation from this representation tractable problems.
Here we will briefly explain our approach.

It is already common practice in network modeling to think of flow behaviors as vectors.
The \caps{IP} protocol number, source and destination \caps{IP} addresses, and \caps{UDP/TCP} source and port numbers that define a packet flow are commonly referred to as a 5-tuple---i.e., a vector in five dimensions.
Other aspects of a flow's behavior can be represented vectorially as well. We call such a mapping of flows into a vector space a \emph{linear representation} if it satisfies the following simple linearity criterion:
\begin{quote}
The sum of the representations of two flows must be another vector that represents the aggregate behavior of the two flows.
\end{quote}
We show that flows can be represented satisfying this criterion. This immediately allows us to perform useful computations on traffic patterns through simple vector addition and scaling.
Finally, we argue that the natural generalization of this approach is to represent an entire network's behavior as a large matrix, with each row representing a single flow.
The matrix representation gives a representation of network behavior that preserves the non-uniformity and non-independence of flow behaviors, but still is highly amenable to analysis using theoretical and numerical linear algebra.
Moreover, we describe how to generate traffic from arbitrary traffic matrices, and demonstrate experimentally that the specific matrix representation we propose preserves the performance characteristics of real traffic.

% TODO: update roadmap paragraph.

@@
